---
title: "EDA.Rmd"
output: html_document
date: "2024-11-11"
---

# Load Data
Set root directory
```{r}
root_dir = paste0(dirname(getwd()), '/')
```

## Combine 2000-2010 Census Data
```{r}

data_path = paste0(root_dir, "data/census_data/00-10_census_data/")

# create list of all the files in the data path that end in a two digit number.csv
file_names = list.files(data_path, pattern = "\\d{2}.csv$", full.names = TRUE)

# Use lapply to read in each file as a data frame
file_data = lapply(file_names, read.csv)

# combine them into a single data frame
combined_census = do.call(rbind, file_data) 

head(combined_census)

write.csv(combined_census, paste0(root_dir, "data/census_data/cc-est2010-alldata.csv"), row.names = FALSE)

census10 = combined_census
```

## Load Other Data
```{r}
census20 = read.csv(paste0(root_dir, 'data/census_data/CC-EST2020-ALLDATA.csv'))
election = read.csv(paste0(root_dir, 'data/election_data/countypres_2000-2020.csv'))
```

```{r}
head(census20)
head(election)
```

# 2000-2010 Census Data Sanity Checks
```{r}
# check the number of rows and columns
dim(census20)
```

```{r}
# Check for duplicated rows
sum(duplicated(census20))
```


## Number of blank cases in each column
```{r}
colSums(is.na(census20)) 
```

There are no NA values, but there are 'X' values in the year 1 data of the Chugach Census Area and Copper River Census Area in Alaska.
This is because, as disclosed in the pdf that accompanies this data, "Effective January 2, 2019, two new Alaska county-equivalent areas, Chugach Census Area and Copper River Census Area, were formed from Valdez-Cordova Census Area, Alaska."
```{r}
# there are no NA values, but there are some 'X' values
# View rows where any column contains the character 'X'
census20[census20$TOT_POP == "X", ]

# for the purposes of exploratory data analysis, I will be removing these census areas
census20 = census20 |>
  filter(CTYNAME != "Chugach Census Area") |>
  filter(CTYNAME != "Copper River Census Area")

```

## Look at total distribution of each column (is it the shape you expect?)
```{r}
colnames = colnames(census20) 
# remove string columns
colnames = setdiff(colnames, c("STNAME","CTYNAME"))


# convert population columns to numeric
census20[, colnames] <- apply(
  census20[, colnames], 
  2, 
  function(x) as.numeric(as.character(x))  # Convert to character before as.numeric for factors
)
# check that there are no NA values
census20[apply(census20, 1, function(x) any(is.na(x))), ]
```

```{r}
hist_no_outliers = function(df, column_name) {
  # Check if the column exists in the dataframe
  if (!(column_name %in% colnames(df))) {
    stop("Column does not exist in the dataframe.")
  }
  
  # Extract the column data
  column_data = df[[column_name]]
  
  # Calculate the IQR
  Q1 = quantile(column_data, 0.25)
  Q3 = quantile(column_data, 0.75)
  IQR_value = IQR(column_data)
  
  # Define the acceptable range for non-outlier values
  lower_bound = Q1 - 1.5 * IQR_value
  upper_bound = Q3 + 1.5 * IQR_value
  
  # Filter out outliers
  df_no_outliers = df[column_data >= lower_bound & column_data <= upper_bound, ]
  
  # Plot histogram without outliers
  hist(df_no_outliers[[column_name]], main = paste("Histogram of", column_name, "(No Outliers)"), 
       xlab = column_name)
}

```

```{r}
# I ran this function for a much larger portion of the columns, but did not include them here so this document would not be too long
hist_no_outliers(census20, "TOT_POP")
hist_no_outliers(census20, "TOT_FEMALE")
hist_no_outliers(census20, "TOT_MALE")

```
## Check for outliers
```{r}
# Because the majority of counties don't have many people and there are a few counties with a lot of people, the box plots make sense
boxplot(census20$TOT_POP, main = "Total Population Boxplot")
boxplot(census20$TOT_MALE, main = "Total Male Population Boxplot")
boxplot(census20$TOT_FEMALE, main = "Total Female Population Boxplot")
```

## Do pop totals add up to total pop
```{r}
population_totals = function(rand_row) {
  # get all the county and year info from that row
  county_data = census20 |>
      filter(STATE == rand_row$STATE) |>
      filter(COUNTY == rand_row$COUNTY) |>
      filter(YEAR == rand_row$YEAR) 
  
  tot_pop = county_data |> 
    filter(AGEGRP == 0) |>
    pull(TOT_POP)
    
  
  # Check total male and female 
  tot_m = county_data |> 
    filter(AGEGRP == 0) |>
    pull(TOT_MALE)
  
  tot_f = county_data |> 
    filter(AGEGRP == 0) |>
    pull(TOT_FEMALE)
  
  if ((tot_m + tot_f) == tot_pop) {
    print("Male and Female populations sum to total")
  } else {
    print("WARNING: Male and Female populations do not sum to the total")
  }
}


# pick random row
rand_row = census20[sample(nrow(census20), 1), ]

population_totals(rand_row)

```






# 2000-2010 Sanity Checks
```{r}
# check the number of rows and columns
# the dimensions check with what I would expect
dim(census10)
```

```{r}
# Check for duplicated rows
sum(duplicated(census10))
```


## Number of blank cases in each column
```{r}
colSums(is.na(census10)) 
```

## Look at total distribution of each column (is it the shape you expect?)
```{r}
colnames = colnames(census10) 
# remove string columns
colnames = setdiff(colnames, c("STNAME","CTYNAME"))

# convert population columns to numeric
census10[, colnames] <- apply(
  census10[, colnames], 
  2, 
  function(x) as.numeric(as.character(x))  # Convert to character before as.numeric for factors
)
# check that there are no NA values
census20[apply(census10, 1, function(x) any(is.na(x))), ]
```


```{r}
# I ran this function for a much larger portion of the columns, but did not include them here so this document would not be too long
hist_no_outliers(census10, "TOT_POP")
hist_no_outliers(census10, "TOT_FEMALE")
hist_no_outliers(census10, "TOT_MALE")

```
## Check for outliers
```{r}
# Because the majority of counties don't have many people and there are a few counties with a lot of people, the box plots make sense
boxplot(census10$TOT_POP, main = "Total Population Boxplot")
boxplot(census10$TOT_MALE, main = "Total Male Population Boxplot")
boxplot(census10$TOT_FEMALE, main = "Total Female Population Boxplot")
```

## Do pop totals add up to total pop
```{r}
# pick random row
rand_row = census10[sample(nrow(census10), 1), ]

population_totals(rand_row)

```


# Election Data Sanity Checks
Right number of rows  and columns
```{r}
# the dimensions check with what I would expect
dim(election)
```
## Check for duplicates 
```{r}
# Check for duplicated rows
sum(duplicated(election))
```


## Number of blank cases in each column
```{r}
colSums(is.na(election)) 
# There are 52 NA values in the county_fips

election |> filter(is.na(county_fips))
# these are for maine, conneticut, and rhode island, where there are federal precincts, statewide write ins, and overseas voters
# these are something to be aware of when I go through each state
```

## Look at total distribution of each column (is it the shape you expect?)
```{r}
histogram_no_outliers = function(col) {
  # Extract the column data
  column_data = df[[column_name]]
  
  # Calculate the IQR
  Q1 = quantile(column_data, 0.25)
  Q3 = quantile(column_data, 0.75)
  IQR_value = IQR(column_data)
  
  # Define the acceptable range for non-outlier values
  lower_bound = Q1 - 1.5 * IQR_value
  upper_bound = Q3 + 1.5 * IQR_value
  
  # Filter out outliers
  df_no_outliers = df[column_data >= lower_bound & column_data <= upper_bound, ]
  
  # Plot histogram without outliers
  hist(df_no_outliers[[column_name]], main = paste("Histogram of", column_name, "(No Outliers)"), 
       xlab = column_name)
}

hist_no_outliers(election, "candidatevotes")
# the large amount of candidate votes clustered around zero makes sense because there are a number of third party candidates who dont get very many votes
hist_no_outliers(election, "totalvotes")

```
## Check for outliers
```{r}
boxplot(election$candidatevotes, main = "Candidate Votes")
boxplot(election$totalvotes, main = "Total Votes")
# the large number of outliers also makes sense in this data set because there will be a portion of counties with large number of votes, while the rest have smaller numbers of votes (the number of votes is proportional to population)
```

## Do candidate totals add up?
```{r}
# pick random row
rand_row = election[sample(nrow(election), 1), ]


# get all the county and year info from that row
county_data = election |>
    filter(state == rand_row$state) |>
    filter(county_name == rand_row$county_name) |>
    filter(year == rand_row$year) 

tot_votes = election |> 
  filter(AGEGRP == 0) |>
  pull(TOT_POP)
  


if (cand_sum == tot_votes) {
  print("Candidate votes sum to total")
} else {
  print("WARNING: Candidate votes do not sum to the total")
}


```
