---
title: "analysis"
output: html_document
---


```{r}
options(scipen = 5)
library(tidyr)
library(stringr)
library(tidyverse)
library(dplyr)
```

```{r}
root_dir = paste0(dirname(getwd()), '/')
flipped20 = read.csv(paste0(root_dir, 'data/flipped20.csv'))
```



```{r}

win_margins = flipped20 |>
  distinct(percent_rep) |>
  pull(percent_rep)

hist(win_margins * 100, 
     main = "Distribution of Percent Democratic Votes \n in Counties that Flipped Party between 2016 and 2020", 
     xlab = "Percent Republican (%)", 
     col = "lightblue", 
     border = "black")
```

```{r}

# analyze flipped county demographics
demographic_cols = c("TOT_MALE","TOT_FEMALE","WA_MALE","WA_FEMALE",
                       "BA_MALE","BA_FEMALE","IA_MALE","IA_FEMALE","AA_MALE","AA_FEMALE",
                       "TOM_MALE","TOM_FEMALE","H_MALE","H_FEMALE",
                       "MINORITY_MALE","MINORITY_FEMALE")
  
flipped20_perc = flipped20 |> 
  mutate(across(
    all_of(demographic_cols), 
    ~ .x / TOT_POP,  # Note the correct placement of the tilde (~)
    .names = "PERC_{col}"  # Creates new column names with a "Perc_" prefix
  )) |>
  select(!all_of(demographic_cols))
  
perc_dem_cols = paste0("PERC_",demographic_cols)

```


```{r}
library(dplyr)
library(ggplot2)
significant_dem_changes <- function(data, demographic_cols) { 
  
  # Calculate changes between 2016 and 2020
  changes = data |>
    group_by(STNAME, CTYNAME, county_fips, age) |>
    reframe(across(all_of(c(demographic_cols, "percent_rep")), ~ diff(.x[order(!!sym("YEAR"))]),
      .names = "Change_{col}"), 
      ) |>
    pivot_longer(cols = starts_with("Change_") & !matches("percent_rep|TOT_POP"), names_to = "Demographic", values_to = "Demographic_Change_Percent") |>
    mutate(Demographic = gsub("Change_", "", Demographic))|>
    mutate(Demographic_Change_Percent = as.numeric(Demographic_Change_Percent)) |>
    arrange(desc(abs(Change_percent_rep)), desc(abs(Demographic_Change_Percent)))
  return(changes)
  
}
 label_mapping <- c(
  "PERC_TOT_MALE" = "Total Male Population",
  "PERC_TOT_FEMALE" = "Total Female Population",
  "PERC_WA_MALE" = "White Male Population",
  "PERC_WA_FEMALE" = "White Female Population",
  "PERC_BA_MALE" = "Black Male Population",
  "PERC_BA_FEMALE" = "Black Female Population",
  "PERC_IA_MALE" = "Indigenous Male Population",
  "PERC_IA_FEMALE" = "Indigenous Female Population",
  "PERC_H_MALE" = "Hispanic Male Population",
  "PERC_H_FEMALE" = "Hispanic Female Population",
  "PERC_MINORITY_MALE" = "Minority Male Population",
  "PERC_MINORITY_FEMALE" = "Minority Female Population",
  "PERC_AA_MALE" = "Asian Male Population",
  "PERC_AA_FEMALE" = "Asian Female Population",
  "PERC_TOM_MALE" = "Two or More Races Male Population",
  "PERC_TOM_FEMALE" = "Two or More Races Female Population"
  )
  
total_changes = significant_dem_changes(data = flipped20_perc, demographic_cols = perc_dem_cols) |>
    mutate(DemographicLabel = label_mapping[Demographic])

win_margins = total_changes |>
  group_by(county_fips, Change_percent_rep) |>
  distinct(county_fips, Change_percent_rep) 

ggplot(win_margins, aes(x = Change_percent_rep * 100)) +
  geom_histogram(breaks = seq(floor(-10), ceiling(25), by = 5),
                 color = "black", 
                 fill = "lightblue") +
  labs(
    title = "Histogram of Change in Percent Republican Votes \n in Counties that Flipped Party between 2016 and 2020",
    x = "Change Percent Republican (%)",
    y = "Frequency"
  ) +
  theme_minimal()

output_dir = paste0(dirname(getwd()), '/output/flipped_counties_all/')
ggplot(win_margins, aes(x = Change_percent_rep * 100)) +
  geom_histogram(breaks = seq(floor(-10), ceiling(25), by = 1),
                 color = "black", 
                 fill = "lightblue") +
  labs(
    title = "Histogram of Change in Percent Republican Votes \n in Counties that Flipped Party between 2016 and 2020",
    x = "Change Percent Republican (%)",
    y = "Frequency"
  ) +
  theme_minimal()
ggsave(paste0(output_dir, 'hist.png'))



```


```{r}
demo_change_avg <- function(data, filename, vals, labs, description) {
  output_dir = paste0(dirname(getwd()), paste0('/output/flipped_counties_all/', filename))

  data = data |>
  group_by(age, Demographic, DemographicLabel) |>
  summarise(Demographic_Change_Percent = mean(Demographic_Change_Percent),
            .groups = "drop") |>
  arrange(desc(abs(Demographic_Change_Percent))) 


  all_bar = data |>
    filter(!(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE"))) |>
    mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
    select(group_label, Demographic_Change_Percent) |>
    ggplot(aes(x = reorder(group_label, Demographic_Change_Percent), y = Demographic_Change_Percent * 100, fill = Demographic_Change_Percent > 0)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = vals,
                    labels = labs)+
    labs(
      title =  paste("All Average Demographic Changes \n in Counties that Flipped Party and Shifted",description), 
      x = "Demographic (by age group, race/ethnicity, and sex)",
      y = "Percent Change (%)",
      fill = "Populationn Shift"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'all_bar.png'), plot = all_bar, height = 15)
  
  
  # Largest Demographic Changes
  biggest_change = data |> 
    filter(abs(Demographic_Change_Percent) >= .005) |>
    mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
    filter(!(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE"))) |>
    select(group_label, Demographic_Change_Percent) 


 
  biggest_bar_colors = biggest_change |> ggplot(aes(x = reorder(group_label, Demographic_Change_Percent), y = Demographic_Change_Percent * 100, fill = Demographic_Change_Percent > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = vals,
                    labels = labs) +
  labs(
    title =  paste("Largest Average Demographic Changes in Counties \n that Flipped Party and Shifted", description), 
    x = "Demographic (by age group, race/ethnicity, and sex)",
    y = "Percent Change (%)",
    fill = "Population Shift"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'biggest_bar.png'), plot = biggest_bar_colors)
  
  # remove age
  no_age_bar_color = data |>
  filter(age == "total") |>
  filter(!(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE"))) |>
  select(-age) |> ggplot(aes(x = reorder(DemographicLabel, Demographic_Change_Percent), y = Demographic_Change_Percent * 100, fill = Demographic_Change_Percent >   0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = vals,
                    labels = labs) +
  labs(
    title =  paste("Average Demographic Changes \n in Counties that Flipped Party and Shifted", description), 
    x = "Demographic (by race/ethnicity and sex)",
    y = "Percent Change (%)",
    fill = "Population Shift"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'no_age_bar.png'), plot = no_age_bar_color)
  
  # age and gender
  age_gender_bar_color = data |>
  filter(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE"))|>
  filter(age != 'total') |>
  mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
  ggplot(aes(x = reorder(group_label, Demographic_Change_Percent), y = Demographic_Change_Percent * 100, fill = Demographic_Change_Percent > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = vals,
                    labels = labs) +
  labs(
    title =  paste("Average Demographic Changes \n in Counties that Flipped Party and Shifted", description), 
    x = "Demographic (by age group and gender)",
    y = "Percent Change (%)",
    fill = "Population Shift"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'age_gender_bar.png'), plot = age_gender_bar_color)
  
  
  # just total female male
  gender = data |>
    filter(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE")) |>
    filter(age == 'total') |>
    mutate(DemographicLabel = if_else(Demographic == "PERC_TOT_MALE", "Total Male Population", "Total Female Population"))|>
  ggplot(aes(x = reorder(DemographicLabel, Demographic_Change_Percent), y = Demographic_Change_Percent * 100, fill = Demographic_Change_Percent > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = vals,
                    labels = labs) +
  labs(
    title =  paste("Average Demographic Changes \n in Counties that Flipped Party and Shifted", description), 
    x = "Demographic (by gender)",
    y = "Percent Change (%)",
    fill = "Population Shift"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'gender_bar.png'), plot = age_gender_bar_color)
  
  
  
  list(data = data,
       all_bar = all_bar,
       biggest_bar = biggest_bar_colors,
       no_age_bar = no_age_bar_color,
       age_gender_bar = age_gender_bar_color,
       gender_bar = gender
       )

}
```

#### dem 2-6
```{r}
dem_2_6 = total_changes |>
  filter((Change_percent_rep < -.02) & (Change_percent_rep >= -.06))

#set colors:
vals = c("TRUE" = "lightskyblue1", "FALSE" = "lightskyblue3")
labs = c("Population Decrease", "Population Increase")
description = "2-6% fewer Republican Party Votes in 2020 than 2016"
dem_2_6_results = demo_change_avg(dem_2_6, 'dem_2_6/', vals, labs, description)

print(dem_2_6_results$data)
# print(dem_2_6_results$all_bar) # its too big to print nicely
print(dem_2_6_results$biggest_bar)
print(dem_2_6_results$no_age_bar)
print(dem_2_6_results$age_gender_bar)
print(dem_2_6_results$gender_bar)
```

#### rep 2-3
```{r}
rep_2_3 = total_changes |>
  filter((Change_percent_rep >= .02) & (Change_percent_rep <= .03))

#set colors:
vals = c("TRUE" = "indianred3", "FALSE" = "indianred4")
labs = c("Population Decrease", "Population Increase")
description = "2-3% more Republican Party Votes in 2020 than 2016"
rep_2_3_results = demo_change_avg(rep_2_3, 'rep_2_3/', vals, labs, description)

print(rep_2_3_results$data)
# print(rep_2_3_results$all_bar) # its too big to print nicely
print(rep_2_3_results$biggest_bar)
print(rep_2_3_results$no_age_bar)
print(rep_2_3_results$age_gender_bar)
print(rep_2_3_results$gender_bar)
```
#### rep 9-11
```{r}
rep_9_10 = total_changes |>
  filter((Change_percent_rep >= .09) & (Change_percent_rep <= .11))

#set colors:
vals = c("TRUE" = "indianred3", "FALSE" = "indianred4")
labs = c("Population Decrease", "Population Increase")
description = "9-11% more Republican Party Votes in 2020 than 2016"
rep_9_10_results = demo_change_avg(rep_9_10, 'rep_9_10/', vals, labs, description)

print(rep_9_10_results$data)
# print(rep_9_10_results$all_bar) # its too big to print nicely
print(rep_9_10_results$biggest_bar)
print(rep_9_10_results$no_age_bar)
print(rep_9_10_results$age_gender_bar)
print(rep_9_10_results$gender_bar)
```

#### rep 19-21
```{r}
rep_19_21 = total_changes |>
  filter((Change_percent_rep >= .19) & (Change_percent_rep <= .21))

#set colors:
vals = c("TRUE" = "indianred3", "FALSE" = "indianred4")
labs = c("Population Decrease", "Population Increase")
description = "19-21% more Republican Party Votes in 2020 than 2016"
rep_19_21_results = demo_change_avg(rep_19_21, 'rep_19_21/', vals, labs, description)

print(rep_19_21_results$data)
# print(rep_19_21_results$all_bar) # its too big to print nicely
print(rep_19_21_results$biggest_bar)
print(rep_19_21_results$no_age_bar)
print(rep_19_21_results$age_gender_bar)
print(rep_19_21_results$gender_bar)
```



Demographic Change by Republican Change is a ratio that quantifies how much a particular demographic's change corresponds to changes in Republican voting behavior, on average, for each group.
A positive value indicates that the demographic change and the Republican vote change are generally aligned in the same direction (e.g., both increasing or both decreasing).
A negative value indicates that the demographic change and the Republican vote change are generally in opposite directions (e.g., one increases while the other decreases).
```{r}
# Example of ratio
example = total_changes |>
  filter(Demographic == "PERC_WA_FEMALE") |>
  mutate(dem_change_by_rep_change = Demographic_Change_Percent / Change_percent_rep) |>
  group_by(age, Demographic, DemographicLabel) |>
  summarise(dem_change_by_rep_change = mean(dem_change_by_rep_change),
            .groups = "drop") |>
   arrange(desc(abs(dem_change_by_rep_change)))
  

```


```{r}

changes = total_changes |> 
   mutate(dem_change_by_rep_change = Demographic_Change_Percent / Change_percent_rep)  |>
  group_by(age, Demographic, DemographicLabel) |>
  summarise(dem_change_by_rep_change = mean(dem_change_by_rep_change),
            .groups = "drop") |>
  arrange(desc(abs(dem_change_by_rep_change)))
```

```{r}
vals = c("TRUE" = "indianred4", "FALSE" = "steelblue4")
labs = c("Democrat", "Republican")

changes |> filter(!(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE"))) |>
    filter(age != 'total') |>
    mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
    select(group_label, dem_change_by_rep_change) |>
    ggplot(aes(x = reorder(group_label, dem_change_by_rep_change), y = dem_change_by_rep_change, fill = dem_change_by_rep_change > 0 )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = vals,
                    labels = labs) +
    labs(
      title = "Correlation of Demographic Change to Increased Republican Votes", 
      x = "Demographic (by age group, race/ethnicity, and sex)",
      y = "Correlation",
      fill = "Party Favorability"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'all.png'), height = 15)
  
changes |>
    filter(age == 'total') |>
    mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
    select(group_label, dem_change_by_rep_change) |>
    ggplot(aes(x = reorder(group_label, dem_change_by_rep_change), y = dem_change_by_rep_change, fill = dem_change_by_rep_change > 0 )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = vals,
                    labels = labs) +
    labs(
      title = "Correlation of Demographic Change to Increased Republican Votes", 
      x = "Demographic (by race/ethnicity and sex)",
      y = "Correlation",
      fill = "Party Favorability"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'totals.png'))
  
  
changes |>
    filter(age == 'total') |>
    filter(Demographic %in% c("PERC_TOT_MALE", "PERC_TOT_FEMALE")) |>
    mutate(group_label = paste(str_to_title(age), DemographicLabel)) |>
    select(group_label, dem_change_by_rep_change) |>
    ggplot(aes(x = reorder(group_label, dem_change_by_rep_change), y = dem_change_by_rep_change, fill = dem_change_by_rep_change > 0 )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = vals,
                    labels = labs) +
    labs(
      title = "Correlation of Demographic Change to Increased Republican Votes", 
      x = "Demographic (by sex)",
      y = "Correlation",
      fill = "Party Favorability"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5) ) # Center-align the title 
  ggsave(paste0(output_dir,'by_sex.png'))

```


